# Names and values

**Learning objectives:**

- test your knowledge
- identify objects in R (size, id, ...)

## Quiz

1. How do I create a new column called "3" that contains the sum of 1 and 2?
```{r}
df <- data.frame(runif(3), runif(3))
names(df) <- c(1, 2)
```

2. How much memory does y occupy?
hint: use `library(lobstr)` 
```{r}
x <- runif(1e6)
y <- list(x, x, x)
```

3. On which line does a get copied in the following example?
```{r}
a <- c(1, 5, 3, 2)
b <- a
b[[1]] <- 10
```



## Object’s identifier

```{r}
x <- c(1, 2, 3)
obj_addr(x)
```


## Exercises

Do they all point to the same underlying function object? hint: `lobstr::obj_addr()`
```{r}
mean
base::mean
get("mean")
evalq(mean)
match.fun("mean")
```


## Meeting Videos

### Cohort 1

(no video recorded)

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/pCiNj2JRK50")`

### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/-bEXdOoxO_E")`

### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/gcVU_F-L6zY")`

### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/aqcvKox9V0Q")`

### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/O4Oo_qO7SIY")`

<details>
<summary> Meeting chat log </summary>

```
00:16:57	Federica Gazzelloni:	cohort 2 video: https://www.youtube.com/watch?v=pCiNj2JRK50
00:18:39	Federica Gazzelloni:	cohort 2 presentation: https://r4ds.github.io/bookclub-Advanced_R/Presentations/Week02/Cohort2_America/Chapter2Slides.html#1
00:40:24	Arthur Shaw:	Just the opposite, Ryan. Very clear presentation!
00:51:54	Trevin:	parquet?
00:53:00	Arthur Shaw:	We may all be right. {arrow} looks to deal with feather and parquet files: https://arrow.apache.org/docs/r/
01:00:04	Arthur Shaw:	Some questions for future meetings. (1) I find Ryan's use of slides hugely effective in conveying information. Would it be OK if future sessions (optionally) used slides? If so, should/could we commit slides to some folder on the repo? (2) I think reusing the images from Hadley's books really helps understanding and discussion. Is that OK to do? Here I'm thinking about copyright concerns. (If possible, I would rather not redraw variants of Hadley's images.)
01:01:35	Federica Gazzelloni:	It's all ok, you can use past presentation, you don't need to push them to the repo, you can use the images from the book
01:07:19	Federica Gazzelloni:	Can I use: gc(reset = TRUE) safely?
```
</details>
