# S3

## Basics

- Has class
- Uses a generic function to decide on method
  - method = implementation for a specific class
  - dispatch = process of searching for right method

## Classes

### Theory

What is class?

  - No formal definition in S3
  - Simply set class attribute

How to set class?

  - At time of object creation
  - After object creation
  
```{r}
# at time of object creation
x <- structure(list(), class = "my_class")

# after object creation
x <- list()
class(x) <- "my_class"
```

Some advice on style:

  - Rules: Can be any string
  - Advice: Consider using/including package name to avoid collision with name of another class (e.g., `blob`, which defines a single class; haven has `labelled` and `haven_labelled`)
  - Convention: letters and `_`; avoid `.` since it might be confused as separator between generic and class name

### Practice

How to compose a class in practice?

- **Constructor**, which helps the developer create new object of target class. Provide always.
- **Validator**, which checks that values in constructor are valid. May not be necessary for simple classes.
- **Helper**, which helps users create new objects of target class. May be relevant only for user-facing classes.

### Constructors

Help developers construct an object of the target class:

```{r}
new_difftime <- function(x = double(), units = "secs") {
  # check inputs
  # issue generic system error if unexpected type or value
  stopifnot(is.double(x))
  units <- match.arg(units, c("secs", "mins", "hours", "days", "weeks"))

  # construct instance of target class
  structure(x,
    class = "difftime",
    units = units
  )
}
```

### Validators

Contrast a constructor, aimed at quickly creating instances of a class, which only checks type of inputs ...

```{r}
new_factor <- function(x = integer(), levels = character()) {
  stopifnot(is.integer(x))
  stopifnot(is.character(levels))

  structure(
    x,
    levels = levels,
    class = "factor"
  )
}

# error messages are for system default and developer-facing
new_factor(1:5, "a")
```


... with a validator, aimed at emitting errors if inputs pose problems, which makes more expensive checks

```{r}
validate_factor <- function(x) {
  values <- unclass(x)
  levels <- attr(x, "levels")

  if (!all(!is.na(values) & values > 0)) {
    stop(
      "All `x` values must be non-missing and greater than zero",
      call. = FALSE
    )
  }

  if (length(levels) < max(values)) {
    stop(
      "There must be at least as many `levels` as possible values in `x`",
      call. = FALSE
    )
  }

  x
}

# error messages are informative and user-facing
validate_factor(new_factor(1:5, "a"))
```

### Helpers

Some desired virtues:

- Have the same name as the class
- Call the constructor and validator, if the latter exists.
- Issue error informative, user-facing error messages
- Adopt thoughtful/useful defaults or type conversion

## Generics and methods

### Generic functions

- Consist of a call to `UseMethod()`
- Pass arguments from the generic to the dispatched method "auto-magically"

```{r}
my_new_generic <- function(x) {
  UseMethod("my_new_generic")
}
```

### Method dispatch

- `UseMethod()` creates a vector of method names
- Dispatch 
  - Examines all methods in the vector
  - Selects a method

```{r}
x <- Sys.Date()
sloop::s3_dispatch(print(x))
```

### Finding methods

While `sloop::s3_dispatch()` gives the specific method selected for a specific call, on can see the methods defined:

- For a generic
```{r}
sloop::s3_methods_generic("mean")
```
- For a class
```{r}
sloop::s3_methods_class("ordered")
```

### Creating methods

Two rules:

- Only write a method if you own the generic. Otherwise, bad manners.
- Method must have same arguments as its generic--with one important exception: `...`

### Examples caught in the wild

- [`haven::zap_label`](https://github.com/tidyverse/haven/blob/main/R/zap_label.R), which removes column labels
- [`dplyr::mutate`](https://github.com/tidyverse/dplyr/blob/main/R/mutate.R)
- [`tidyr::pivot_longer`](https://github.com/tidyverse/tidyr/blob/main/R/pivot-long.R)

## Inheritance

Three ideas:

1. Class is a vector of classes
```{r}
class(ordered("x"))
class(Sys.time())
```
2. Dispatch moves through class vector until it finds a defined method
```{r}
sloop::s3_dispatch(print(ordered("x")))
```
3. Method can delegate to another method via `NextMethod()`, which is indicated by `<-` as below:
```{r}
sloop::s3_dispatch(ordered("x")[1])
```

### NextMethod()

Consider `secret` class that masks each character of the input with `x` in output

```{r}
new_secret <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "secret")
}

print.secret <- function(x, ...) {
  print(strrep("x", nchar(x)))
  invisible(x)
}

x <- new_secret(c(15, 1, 456))
x
```

Notice that the `[` method is problematic in that it does not preserve the `secret` class

```{r}
sloop::s3_dispatch(x[1])
```

Fix this with a `[.secret` method:

```{r}
`[.secret` <- function(x, i) {
  # first, dispatch to `[`
  # then, coerce subset value to `secret` class
  new_secret(NextMethod())
}
```

Notice that `[.secret` is selected for dispatch, but that the method delegates to the internal `[`

```{r}
sloop::s3_dispatch(x[1])
```

### Allowing subclassing

Continue the example above to have a `supersecret` subclass that hides even the number of characters in the input (e.g., `123` -> `xxxxx`, 12345678 -> `xxxxx`, 1 -> `xxxxx`).

To allow for this subclass, the constructor function needs to include two additional arguments:

- `...` for passing an arbitrary set of arguments to different subclasses
- `class` for defining the subclass

```{r}
new_secret <- function(x, ..., class = character()) {
  stopifnot(is.double(x))

  structure(
    x,
    ...,
    class = c(class, "secret")
  )
}
```

To create the subclass, simply invoke the parent class constructor inside of the subclass constructor:

```{r}
new_supersecret <- function(x) {
  new_secret(x, class = "supersecret")
}

print.supersecret <- function(x, ...) {
  print(rep("xxxxx", length(x)))
  invisible(x)
}
```

But this means the subclass inherits all parent methods and needs to overwrite all parent methods with subclass methods that return the sublclass rather than the parent class.

There's no easy solution to this problem in base R.

There is a solution in the vectors package: `vctrs::vec_restore()`

<!-- TODO: read docs/vignettes to be able to summarize how this works -->

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/Fy3JF5Em6qY")`

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/9GkgNC15EAw")`

### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/q7lFXSLdC1g")`

`r knitr::include_url("https://www.youtube.com/embed/2rHS_urTGFg")`

### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/4la5adcWwKE")`

`r knitr::include_url("https://www.youtube.com/embed/eTCT2O58GYM")`

### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/NeHtEGab1Og")`

### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
