```{r, echo= FALSE, message=FALSE}
library(rlang)
library(purrr)
```


# Quasiquotation

**Learning objectives:**

- What quasiquotation means
- Why it's important
- Learn some practical uses

## Introduction

- Three pillars of *tidy* evaluation
   1. Quasiquotation
   2. Quosures (chapter 20)
   3. Data masks (Chapter 20)

- Quasiquotation = quotation + unquotation:
   - **Quote.** Capture unevaluated expression ...("defuse")
   - **Unquote.** Except for selected parts which we do want to evaluate! ("inject")
   
- Functions that use these features are said to use Non-standard evaluation (NSE)

- Note: related to Lisp macros, and also exists in other languages with Lisp heritage, e.g. Julia

## Motivation


Simple *concrete* example:

`Cement` is a function that works like `paste` but doesn't need need quotes:

```{r}
cement <- function(...) {
  args <- ensyms(...)
  paste(purrr::map(args, as_string), collapse = " ")
}

cement(Good, morning, Hadley)
```

What if we wanted to use variables ?   This is where 'unquoting' comes in!

```{r}
name = "Bob"
cement(Good, afternoon, !!name)
```


 
## Nonstandard evaluation {-}

* Functions like `dplyr::filter` use nonstandard evaluation,  and quote some of their arguments to help make code more *tidy*.

```{r}
#| eval: FALSE
# `cyl` is written as a bare name--a symbol defined in the global environment
# but `cyl` only exists in the data frame "environment"
# so, `{dplyr}` quotes the argument
dplyr::filter(mtcars, cyl == 4)
```
 
* You often can detect this if the argument wouldn't work in isolation, for example:

```{r, eval = FALSE}
library(MASS) # this is fine
MASS 
#> Error: object MASS not found
```

and 

```{r, eval = FALSE}
cyl 
#> Error: object 'cyl' not found
```


## Quote

- Expression

```{r}
# for interactive use
rlang::expr(x+y)

# enexpr works on function arguments (looks at internal promise object)  
f2 <- function(x) rlang::enexpr(x)
f2(a + b + c)
```
- To capture multiple arguments, use `enexprs()`

```{r}
f <- function(...) enexprs(...)
f(x=1, y= 10 *z)
```


- For symbols, there is `ensym` and `ensyms` which check that the argument is a symbol or string.

## Base R method {-}

* Base R methods do not support unquoting.

* Base R equivalent of `expr` is `quote`  

* Base R equivalent of `enexpr` is `substitute` (note that `enexpr` uses `substitute`!)

```{r, eval = FALSE}
enexpr
#>function (arg) 
#>{
#>    .Call(ffi_enexpr, substitute(arg), parent.frame())
#>}
```


* `bquote()` provides a limited form of quasiquotation, see section 19.5

* `~`, the formula, is a quoting function, discussed in Section 20.3.4

## Unquote

- Unquoting allows you to 'splice' together ASTs 

- One argument
```{r}
# quote `-1` as `x`
x <- rlang::expr(-1)
# unquote `x` to substitute its unquoted value
# use bang-bang operator
res = rlang::expr(f(!!x, y))
print(res)
lobstr::ast(!!res)
```

- If the right-hand side of `!!` is a function call, it will evalute the function and insert the results.

```{r}
mean_rm <- function(var) {
  var <- ensym(var)
  expr(mean(!!var, na.rm = TRUE))
}
expr(!!mean_rm(x) + !!mean_rm(y))
#> mean(x, na.rm = TRUE) + mean(y, na.rm = TRUE)
```



- Multiple arguments, use `!!!`

```{r}
xs <- rlang::exprs(1, a, -b)
# unquote multiple arguments
# use bang-bang-bang operator
res=expr(f(!!!xs, y))
res
```
```{r}
lobstr::ast(!!res)
```

## ... (dot-dot-dot)

* !!! is also useful in other places where you have a list of expressions you want to insert into a call. 

* Two motivating examples:

   * list of dataframes you want to `rbind`  (a list of arbitrary length)

```{r}
dfs <- list(
  a = data.frame(x = 1, y = 2),
  b = data.frame(x = 3, y = 4)
)
``` 

  * How to supply an argument name indirectly?
  
```{r}
var <- "x"
val <- c(4, 3, 9)
```
  
  
* For the first one, we can use unquote in `dplyr::bind_rows``

```{r}
dplyr::bind_rows(!!!dfs)
```

This is known 'splatting' in some other langauges (Ruby, Go, Julia).  Python calls this argument unpacking (`**kwarg`)

* For the second we need to unquote the left side of an `=`. Tidy  eval lets us do this with a special `:=`

```{r}
tibble::tibble(!!var := val)
```

* Functions that have these capabilities are said to have *tidy dots* (or *dynamic dots*). To get this capability in your own functions, use `list2`!

## Example of `list2()` {-}

```{r}
set_attr <- function(.x, ...) {
  attr <- rlang::list2(...)
  attributes(.x) <- attr
  .x
}

attrs <- list(x = 1, y = 2)
attr_name <- "z"

1:10 %>%
  set_attr(w = 0, !!!attrs, !!attr_name := 3) %>% 
  str()
```
### Exercise from 19.6.5 {-}

What is the problem here?

```{r, eval=FALSE}
set_attr <- function(x, ...) {
  attr <- rlang::list2(...)
  attributes(x) <- attr
  x
}
set_attr(1:10, x = 10)
#> Error in attributes(x) <- attr : attributes must be named
```

## Exec {-}

What about existing functions that don't support tidy dots?  Use `exec`

```{r}
arg_name  <- "na.rm"
arg_val <- TRUE
exec("mean", 1:10, !!arg_name := arg_val)
```

Note that you do not unquote arg_val.
 
Also `exec` is useful for mapping over a list of functions:

```{r}
x <- c(runif(10), NA)
funs <- c("mean", "median", "sd")
purrr::map_dbl(funs, exec, x, na.rm = TRUE)
```



## dots_list {-}

- `list2()` is a wrapper around `dots_list` with the most common defaults:

   - `.ignore_empty` : Ignores any empty arguments, lets you use trailing commas in a list
   - `.homonyms` : controls what happens when multiple arguments use the same name, `list2()` uses default of `keep`
   -  `.preserve_empty` controls what do so with empty arguments if they are not ignored.
   
   
##  Base R `do.call` {-}

`do.call(what, args)` . `what` is a function to call, `args` is a list of arguments to pass to the function.

```{r}
do.call("rbind", dfs)
```
 

### Exercise 19.5.5 #1 {-}

One way to implement `exec` is shown here: Describe how it works. What are the key ideas?

```{r}
exec_ <- function(f, ..., .env = caller_env()){
  args <- list2(...)
  do.call(f, args, envir  = .env)
}
```

## Map-reduce example {-}

Function that will return an expression corresponding to a linear model.

```{r}
linear <- function(var, val) {
  
  # capture variable as a symbol
  var <- ensym(var)
  
  # Create a list of symbols of the form var[[1]], var[[2], etc]
  coef_name <- map(seq_along(val[-1]), ~ expr((!!var)[[!!.x]]))

  # map over the coefficients and the names to create the terms
  summands <- map2(val[-1], coef_name, ~ expr((!!.x * !!.y)))
  
  # Dont forget the intercept
  summands <- c(val[[1]], summands)

  # Reduce!
  reduce(summands, ~ expr(!!.x + !!.y))
}

linear(x, c(10, 5, -4))
#> 10 + (5 * x[[1L]]) + (-4 * x[[2L]])
```


## Creating functions example {-}

* `rlang::new_function()` creates a function from its three components and supports tidy evaluation

* Alternative to function factories.

Example:
```{r}
power <- function(exponent) {
  new_function(
    exprs(x = ), 
    expr({
      x ^ !!exponent
    }), 
    caller_env()
  )
}
power(0.5)
 
```

Another example, is `graphics::curve` that allows you to plot an expression without creating a function. It could be implemented like this:

```{r}
curve2 <- function(expr, xlim = c(0, 1), n = 100) {
  expr <- enexpr(expr)
  f <- new_function(exprs(x = ), expr)
  
  x <- seq(xlim[1], xlim[2], length = n)
  y <- f(x)

  plot(x, y, type = "l", ylab = expr_text(expr))
}
curve2(sin(exp(4 * x)), n = 1000)
```

 
## Summary {-}

* In this chapter we dove into non-standard evaluation with quasiquotation

* Quasiquotation is useful on its own but in the next chapter we will look at the `quosures` and `data masks` to unleash the full power of *tidy evaluation*!



## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/tbByqsRRvdE")`

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/IXE21pR8EJ0")`

### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/gxSpz6IePLg")`

### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/aniKrZrr4aU")`

### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/klcpEb5ZBSM")`

### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/OBodjc80y-E")`

<details>
<summary> Meeting chat log </summary>

```
01:02:07	Trevin:	Yeah, that was a great workshop
01:02:18	Trevin:	Glad they posted the resources online
01:06:39	Trevin:	Thank you!
```
</details>

### Cohort 7

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>

<summary>Meeting chat log</summary>
```
LOG
```
</details>
