# Quasiquotation

**Learning objectives:**

- What quasiquotation is
- Why it might be useful
- How it works
- What are some common use cases

## What is quasiquotation?
 
- **Quote.** Stop evaluation of an expression--typically, a function argument. (In `{rlang}` jargon, defuse.)
- **Unquote.** Evaluate a quoted expression--typically, inside a function. (In `{rlang}` jargon often, inject)

## Why use quasiquotation?

- **Make functions tidy.** Provide function arguments bare names.
```{r}
#| eval: FALSE
# `cyl` is written as a bare name--a symbol defined in the global environment
# but `cyl` only exists in the data frame "environment"
# so, `{dplyr}` quotes the argument
dplyr::filter(mtcars, cyl == 4)
```
- **Make functions Shiny-friendly.** Provide names of columns as strings.
```{r}
#| eval = FALSE
# imagine that `var` came from `input$var_you_selected`
# this "quoted" column name is transformed into a bare name below
my_mean <- function(df, var) {
  dplyr::mutate(df, mean = mean(.data[[var]], na.rm = TRUE))
}
```
- **Inject user-written into a function's pipeline.** For example: summarize expressions, rename expressions, etc.

## How use quasiquotation?

Let's look at its components:

- Quotate
- Unquote
- Quote and unquote, in one go

## Quote

- Expressions
```{r}
#| label:quote_expressions
#| eval:TRUE

# 1. rlang::expr

# for interactive use
rlang::expr(x+y)

# but not for use in a function
# see below

# 2. rlang::enexpr

# expr doesn't yield desired result
f1 <- function(x) rlang::expr(x)
f1(a + b + c)

# enexpr does, since it maintains reference to environment where x is defined
f2 <- function(x) rlang::enexpr(x)
f2(a + b + c)
```
- Symbols
```{r}
#| label:quote_symbols
#| eval:TRUE

# captures a list of symbols or strings in ...
# returns a list of symbols
f <- function(...) {
  ensyms(...)
}

# case 1: input = symbol
f(x)
# case 2: input = character string
f("x")
```

## Unquote

- One argument
```{r}
#| label:unquote_one_arg
#| eval:TRUE

# quote `-1` as `x`
x <- rlang::expr(-1)
# unquote `x` to substitute its unquoted value
# use bang-bang operator
rlang::expr(f(!!x, y))
```
- Multiple arguments
```{r}
#| label:unquote_mult_args
#| eval:TRUE

# quote multiple args--note the `s`
xs <- rlang::exprs(1, a, -b)
# unquote multiple arguments
# use bang-bang-bang operator
expr(f(!!!xs, y))
```

## Quote and unquote

- **Single argument.** Embrace the simplicity of the embrace operator `{{`
```{r}
#| label: embrace
#| eval: FALSE

# The hard way
my_hard_summarize <- function(data, var) {
  # defuse (aka quote)
  var <- rlang::enquo(var)
  
  # inject (aka unquote)
  dplyr::summarize(data, mean = mean(!!var, na.rm = TRUE))
    
}

# The easy way
my_easy_summarize <- function(data, var) {

  # in one move, quote and unquote `var`
  dplyr::summarize(data, mean = mean({{var}}, na.rm = TRUE))
    
}

```
- **Multiple arguments.** Just pass the (dynamic) dots when you can. Otherwise, quote (e.g., `rlang::enquos`) and splice (`!!!`).
```{r}
#| label: dots

# The easy way for easy cases of simple "forwarding"
my_group_by <- function(.data, ...) {
  .data %>% dplyr::group_by(...)
}

mtcars %>% my_group_by(cyl = cyl * 100, am)

# The harder way remains the solution for non-forwarding use cases
# There is no plural version of the embrace operator `{{`
```



## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/tbByqsRRvdE")`

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/IXE21pR8EJ0")`

### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/gxSpz6IePLg")`

### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/aniKrZrr4aU")`

### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/klcpEb5ZBSM")`

### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/OBodjc80y-E")`

<details>
<summary> Meeting chat log </summary>

```
01:02:07	Trevin:	Yeah, that was a great workshop
01:02:18	Trevin:	Glad they posted the resources online
01:06:39	Trevin:	Thank you!
```
</details>

### Cohort 7

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>

<summary>Meeting chat log</summary>
```
LOG
```
</details>
