---
engine: knitr
title: Conditions
---

## Learning objectives:

- What conditions are
- How to use them

## Introduction

What are conditions? Problems that happen in functions:

- Error
- Warning
- Message

As a function author, one can signal them--that is, say there's a problem.

As a function consumer, one can handle them--for example, react or ignore.

## Signalling conditions

### Types of conditions

Three types of conditions:

- `r emoji::emoji("x")` **Errors.** Problem arose, and the function cannot continue. 
- `r emoji::emoji("warning")` **Warnings.** Problem arose, but the function can continue, if only partially.
- `r emoji::emoji("speech_balloon")` **Messages.** Something happened, and the user should know.

### `r emoji::emoji("x")` Errors

How to throw errors

```{r throwing_errors, error = TRUE}
# with base R
stop("... in the name of love...")

# with rlang
rlang::abort("...before you break my heart...")

# with base R; without call
stop("... think it o-o-over...", call. = FALSE)
```
Composing error messages

- Mechanics.
  - `stop()` pastes together arguments
```{r, error = TRUE}
some_val <- 1
stop("Your value is: ", some_val, call. = FALSE)
```
  - `abort()` requires `{glue}`
```{r, error = TRUE}
some_val <- 1
rlang::abort(glue::glue("Your value is: {some_val}"))
```
- Style. See [here](http://style.tidyverse.org/error-messages.html).

### `r emoji::emoji("warning")` Warnings

May have multiple warnings per call

```{r}
warn <- function() {
  warning("This is your first warning")
  warning("This is your second warning")
  warning("This is your LAST warning")
}
```

Print all warnings once call is complete.

```{r}
warn()
```

Like errors, `warning()` has

- a call argument
- an `{rlang}` analog

```{r}
# base R
# ... with call (implicitly .call = TRUE)
warning("Warning")
# ... with call suppressed
warning("Warning", call. = FALSE)

# rlang
# note: call suppressed by default
rlang::warn("Warning")
```

(Hadley's) advice on usage:

- Err on the side of errors. In other words, error rather than warn.
- But warnings make sense in a few cases:
  - Function is being deprecated. Warn that it is reaching end of life.
  - Function is reasonably sure to recover from issue.

### `r emoji::emoji("speech_balloon")` Messages

Mechanics:

- Issued immediately
- Do not have a call argument

Style:

Messages are best when they inform about:

- Default arguments
- Status updates of for functions used primarily for side-effects (e.g., interaction with web API, file downloaded, etc.)
- Progress of long-running process (in the absence of a status bar).
- Package loading message (e.g., attaching package, objects masked)

## Ignoring conditions

A few ways:

- `try()`
- `suppressWarnings()`
- `suppressMessages()`

### `try()`

What it does:

- Displays error
- But continues execution after error

```{r}
bad_log <- function(x) {
  try(log(x))
  10
}

bad_log("bad")
```

Better ways to react to/recover from errors:

1. Use `tryCatch()` to "catch" the error and perform a different action in the event of an error.
1. Set a default value inside the call. See below.

```{r}
default <- NULL
try(default <- read.csv("possibly-bad-input.csv"), silent = TRUE)
```


### `suppressWarnings()`, `suppressMessages()`

What it does:

- Supresses all warnings (messages)

```{r}
# suppress warnings (from our `warn()` function above)
suppressWarnings(warn())

# suppress messages
many_messages <- function() {
  message("Message 1")
  message("Message 2")
  message("Message 3")
}

suppressMessages(many_messages())
```

## Handling conditions

Every condition has a default behavior:

- `r emoji::emoji("x")` Errors halt execution
- `r emoji::emoji("warning")` Warnings are collected during execution and displayed in bulk after execution
- `r emoji::emoji("speech_balloon")` Messages are displayed immediately

Condition handlers allow one to change that behavior (within the scope of a function).

Two handler functions:

- `tryCatch()`
- `withCallingHandlers()`

```{r, eval=FALSE}
# try to run `code_to_try_to_run`
# if (error) condition is signalled, fun some other code
tryCatch(
  error = function(cnd) {
    # code to run when error is thrown
  },
  code_to_try_to_run
)

# try to `code_to_try_to_run`
# if condition is signalled, run code corresponding to condition type
withCallingHandlers(
  warning = function(cnd) {
    # code to run when warning is signalled
  },
  message = function(cnd) {
    # code to run when message is signalled
  },
  code_to_try_to_run
)
```


### Condition objects

```{r}
# catch a condition
cnd <- rlang::catch_cnd(stop("An error"))
# inspect it
str(cnd)
```

The standard components

- `message`. The error message. To extract it, use `conditionMessage(cnd)`.
- `call`. The function call that triggered the condition. To extract it, use `conditionCall(cnd)`.

But custom conditions may contain other components.

### Exiting handlers

If a condition is signalled, this type of handler controls what code to run before exiting the function call. 

```{r}
f3 <- function(x) {
  tryCatch(
    # if error signalled, return NA
    error = function(cnd) NA,
    # try to run log
    log(x)
  )
}

f3("x")
```

When a condition is signalled, control moves to the handler and never returns to the original code.

```{r}
tryCatch(
  message = function(cnd) "There",
  {
    message("Here")
    stop("This code is never run!")
  }
)
```

The `tryCatch()` exit handler has one final argument: `finally`. This is run regardless of the condition of the original code. This is often used for clean-up.

```{r}
# try to write text to disk
# if an error is signalled--for example, `path` does not exist
# or if no condition is signalled
# that is in both cases, the code block in `finally` is executed
path <- tempfile()
tryCatch(
  {
    writeLines("Hi!", path)
    # ...
  },
  finally = {
    # always run
    unlink(path)
  }
)
```

### Calling handlers

Definition by verbal comparison:

- With exit handlers, code exits the normal flow once a condition is signalled
- With calling handlers, code continues in the normal flow once control is returned by the handler.

Definition by code comparison:

```{r}
# with an exit handler, control moves to the handler once condition signalled and does not move back
tryCatch(
  message = function(cnd) cat("Caught a message!\n"), 
  {
    message("Someone there?")
    message("Why, yes!")
  }
)

# with a calling handler, control moves first to the handler and the moves back to the main code
withCallingHandlers(
  message = function(cnd) cat("Caught a message!\n"), 
  {
    message("Someone there?")
    message("Why, yes!")
  }
)
```

### By default, conditions propagate

Let's suppose that there are nested handlers. If a condition is signalled in the child, it propagates to its parent handler(s).

```{r}
# Bubbles all the way up to default handler which generates the message
withCallingHandlers(
  message = function(cnd) cat("Level 2\n"),
  withCallingHandlers(
    message = function(cnd) cat("Level 1\n"),
    message("Hello")
  )
)

# Bubbles up to tryCatch
tryCatch(
  message = function(cnd) cat("Level 2\n"),
  withCallingHandlers(
    message = function(cnd) cat("Level 1\n"),
    message("Hello")
  )
)
```

### But conditions can be muffled

If one wants to "muffle" the siginal, one needs to use `rlang::cnd_muffle()`

```{r}
# Muffles the default handler which prints the messages
withCallingHandlers(
  message = function(cnd) {
    cat("Level 2\n")
    rlang::cnd_muffle(cnd)
  },
  withCallingHandlers(
    message = function(cnd) cat("Level 1\n"),
    message("Hello")
  )
)

# Muffles level 2 handler and the default handler
withCallingHandlers(
  message = function(cnd) cat("Level 2\n"),
  withCallingHandlers(
    message = function(cnd) {
      cat("Level 1\n")
      rlang::cnd_muffle(cnd)
    },
    message("Hello")
  )
)
```

### Call stacks

Call stacks of exiting and calling handlers differ.

Why? 

> Calling handlers are called in the context of the call that signalled the condition
> exiting handlers are called in the context of the call to tryCatch()

To see this, consider how the call stacks differ for a toy example.

```{r}
# create a function
f <- function() g()
g <- function() h()
h <- function() message

# call stack of calling handlers
withCallingHandlers(f(), message = function(cnd) {
  lobstr::cst()
  rlang::cnd_muffle(cnd)
})

# call stack of exit handlers
tryCatch(f(), message = function(cnd) lobstr::cst())
tryCatch(f(), message = function(cnd) lobstr::cst())
```

## Custom conditions

### Motivation

The `base::log()` function provides a minimal error message.

```{r, error = TRUE}
log(letters)
log(1:10, base = letters)
```

One could make a more informative error message about which argument is problematic.

```{r}
my_log <- function(x, base = exp(1)) {
  if (!is.numeric(x)) {
    rlang::abort(paste0(
      "`x` must be a numeric vector; not ", typeof(x), "."
    ))
  }
  if (!is.numeric(base)) {
    rlang::abort(paste0(
      "`base` must be a numeric vector; not ", typeof(base), "."
    ))
  }

  base::log(x, base = base)
}
```

Consider the difference:

```{r, error = TRUE}
my_log(letters)
my_log(1:10, base = letters)
```


### Signalling

Create a helper function to describe errors:

```{r}
abort_bad_argument <- function(arg, must, not = NULL) {
  msg <- glue::glue("`{arg}` must {must}")
  if (!is.null(not)) {
    not <- typeof(not)
    msg <- glue::glue("{msg}; not {not}.")
  }
  
  rlang::abort(
    "error_bad_argument", # <- this is the (error) class, I believe
    message = msg, 
    arg = arg, 
    must = must, 
    not = not
  )
}
```

Rewrite the log function to use this helper function:

```{r}
my_log <- function(x, base = exp(1)) {
  if (!is.numeric(x)) {
    abort_bad_argument("x", must = "be numeric", not = x)
  }
  if (!is.numeric(base)) {
    abort_bad_argument("base", must = "be numeric", not = base)
  }

  base::log(x, base = base)
}
```

See the result for the end user:

```{r, error = TRUE}
my_log(letters)
my_log(1:10, base = letters)
```

### Handling

Use class of condition object to allow for different handling of different types of errors

```{r, error = TRUE}
tryCatch(
  error_bad_argument = function(cnd) "bad_argument",
  error = function(cnd) "other error",
  my_log("a")
)
```

But note that the first handler that matches any of the signal's class, potentially in a vector of signal classes, will get control. So put the most specific handlers first.

## Applications

See [the sub-section in the book](https://adv-r.hadley.nz/conditions.html#condition-applications) for excellent examples.

## Resources

- Conditions articles in rlang vignettes: 
  - [Including function calls in error messages](https://rlang.r-lib.org/reference/topic-error-call.html)
  - [Including contextual information with error chains](https://rlang.r-lib.org/reference/topic-error-chaining.html)
  - [Formatting messages with cli](https://rlang.r-lib.org/reference/topic-condition-formatting.html)
- [Other resources](https://github.com/rstudio-conf-2022/pkg-dev-masterclass/blob/main/materials/5-error-resources.md) from error message segment of rstudio::conf(2022) workshop "Package Development Masterclass"

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/mwiNe083DLU")`

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/ZFUr7YRSu2o")`

### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/UZhrsVz6wi0")`

`r knitr::include_url("https://www.youtube.com/embed/Wt7p71_BuYY")`

### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/WinIo5mrUZo")`

### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/VFs-2sl5C70")`

### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/VwmrbPUQY1k")`

<details>
<summary> Meeting chat log </summary>

```
00:19:16	Trevin:	https://style.tidyverse.org/error-messages.html
00:20:14	Trevin:	More on errors in the design guide: https://design.tidyverse.org/
01:14:27	Federica Gazzelloni:	more info here: https://colinfay.me/learn-shiny-production/
```
</details>

### Cohort 7

`r knitr::include_url("https://www.youtube.com/embed/t1N6XdidvNo")`

<details>

<summary>Meeting chat log</summary>
```
00:34:09	Ron:	Someone did: https://cran.r-project.org/web/packages/comprehenr/vignettes/Introduction.html
00:47:58	collinberke:	https://purrr.tidyverse.org/reference/safely.html
00:48:24	Ron:	it's a function operator !
00:49:37	Ron:	\(x) length(unique(x) is not too verbose though
00:49:39	Ron:	;)
01:06:50	collinberke:	https://colinfay.me/purrr-mappers/
01:07:45	collinberke:	https://colinfay.me/purrr-web-mining/
```
</details>
