---
engine: knitr
title: S7
---

## Learning objectives

- Recognize the challenges with S3 and S4 that motivated development of S7
- Create S7 classes
- Define custom validators for S7 classes and properties
- Create S7 generics and methods
- Use S7 with S3 and/or S4

```{r}
library(S7)
```

# Motivation for S7

## S7 resolves challenges with S3

- S3 has no formal class definitions
- S3 methods are difficult to find
- S3 properties are in attributes, but `attr()` does fuzzy matching
- S3 method dispatch via `UseMethod()` is fuzzy

> "Now for some obscure details that need to appear somewhere" —`?UseMethod`

- `NextMethod()` depends on what's loaded
- Conversion between S3 classes is fuzzy (*vs* `S7::convert()` generic)

## S7 resolves challenges with S4

- S4's multiple inheritance causes more problems than it solves
- S4's method dispatch is smart but hard to predict (S7 is explicit)
- S4 is clean break from S3, made it hard to switch from S3 to S4
- S4 pretends users can't use `@` to access slots, but they do it anyway

## S7 extends S3 and replaces S4

- Name comes from S3 + S4 = S7, but...
  - S7 objects *are* S3 objects
  - S7 objects *are not* S4 objects
  - But there's some overlap (see last section)

# Classes and objects (and properties)

::: notes
This section combines the [1st section of S7 basics](https://rconsortium.github.io/S7/articles/S7.html#classes-and-objects) with the [Classes and objects vignette](https://rconsortium.github.io/S7/articles/classes-objects.html) 
:::

## Define S7 classes with `S7::new_class()`

```{r}
class_person <- new_class(
  "Person",
  properties = list(name = class_character, age = class_numeric)
)
me <- class_person(name = "Jon", age = 50)
me
S7_inherits(me, class_person)
class(me)
inherits(me, "Person")
```

::: notes
- Vignette shows same name for the class object and the 1st arg, but in practice I find it safer to differentiate them (in case you use the class as a property of another class).
  - Notice that you use `class_character`, not `character`; I think that's connected to the issues I had with name matching
- Equivalent to constructor + validator + helper from S3 chapter
  - We'll see this more explicitly later
- Equivalent to `setClass()` + `new()` + `setValidity()` from S4 chapter
- S3 methods will dispatch properly for this object.
:::

## Access S7 object properties with `@`

```{r}
me@name
me@age
```

::: notes
- Like S3 attributes, but `attr()` is fuzzy, and S3 attributes are less formally defined.
- Like S4 slots, but S4 users are discouraged from using `@` directly, and S4 `@` doesn't trigger validation (see next).
:::

## S7 objects are validated during construction and on property assignment

```{r}
#| error: true
me@age <- "fifty"
us <- class_person(name = 1:2, age = c("fifty", "forty-nine"))
```

## `validator` argument customizes validation

```{r}
#| error: true
class_person <- new_class(
  "Person",
  properties = list(name = class_character, age = class_numeric),
  validator = function(self) {
    if (length(self@name) != length(self@age)) {
      "@name and @age must be the same length"
    }
  }
)
us <- class_person(name = c("Jon", "Leyla"), age = c(50, -5, 49))
me <- class_person("Jon", 50)
me@age <- 50:60
```

::: notes
Can return more than one string.
:::

## Set properties all at once to avoid intermediate invalid states

```{r}
#| error: true
us <- me
us@name <- c("Jon", "Leyla")
# Quick S3 method to demonstrate
c.Person <- function(x, y) {
  props(x) <- list(name = c(x@name, y@name), age = c(x@age, y@age))
  x
}
us <- c(me, class_person("Leyla", 49))
us
```

::: notes
This is also a preview of why S7 is easier to adopt in an S3 world
:::

## `S7::new_property()` defines custom property types

```{r}
#| error: true
prop_positive <- new_property(
  class = class_numeric,
  validator = function(value) {
    if (any(value <= 0)) "must be positive"
  }
)
class_person <- new_class(
  "Person",
  properties = list(name = class_character, age = prop_positive),
  validator = function(self) {
    if (length(self@name) != length(self@age)) {
      "@name and @age must be the same length"
    }
  }
)
us <- class_person(name = c("Jon", "Leyla"), age = c(50, -5))
```

::: notes
Can also use this to set default values
:::

## Properties can be computed

```{r}
class_circle <- new_class(
  "Circle",
  properties = list(
    radius = class_numeric,
    area = new_property(
      class = class_numeric,
      getter = function(self) {
        pi * self@radius^2
      }
    )
  )
)
c1 <- class_circle(radius = 1)
c1@area == pi
c1@radius <- 2
c1@area == 4*pi
```

## Properties can be fully dynamic

```{r}
class_circle2 <- new_class(
  "Circle",
  properties = list(
    radius = class_numeric,
    area = new_property(
      class = class_numeric,
      getter = function(self) pi * self@radius^2,
      setter = function(self, value) {
        if (!length(value)) return(self)
        self@radius <- sqrt(value / pi)
        self
      }
    )
  )
)
c2 <- class_circle2(radius = 1)
c2@area
c2@area <- 4*pi
c2@radius
c2@radius <- 3
c2@area == 9*pi
```

::: notes
- This starts to feel kinda encapsulated like we'll see in R6
- Still works with S3-style multimethods, though
:::

## `constructor` argument customizes object creation

- (it might be nice to have a slide on this but we don't yet)

# Generics and methods

::: notes
This section combines the [2nd and 3rd section of S7 basics](https://rconsortium.github.io/S7/articles/S7.html#generics-and-methods) with the [Generics and methods vignette](https://rconsortium.github.io/S7/articles/generics-methods.html#generic-method-compatibility) 
:::

## Define S7 generics & methods with `S7::new_generic()` and `S7::new_method()`

```{r}
are_old <- new_generic("are_old", "x")
method(are_old, class_person) <- function(x) {
  x@age >= 40 & x@name != "Leyla"
}
us
are_old(us)
```

## S7 method dispatch is similar to S3

```{r}
class_geek <- new_class(
  "Geek",
  parent = class_person
)
me <- class_geek("Jon", 50)
are_old(me) # Dispatches on "Person"
```

## S7 method chaining is explicit

```{r}
method(are_old, class_geek) <- function(x) {
  cat("Checking if geeks are old...\n")
  are_old(super(x, class_person)) # Must specify which parent class to use
}
are_old(me)
```

## S7 generics allow for multiple dispatch

```{r}
combine <- new_generic("combine", c("x", "y"))
method(combine, list(class_person, class_person)) <- function(x, y) {
  class_person(
    name = c(x@name, y@name),
    age = c(x@age, y@age)
  )
}
hw <- class_geek("Hadley", 46)
combine(us, hw)
combine(me, hw)
method(combine, list(class_geek, class_geek)) <- function(x, y) {
  combined_person <- combine(
    super(x, class_person),
    super(y, class_person)
  )
  class_geek(
    name = combined_person@name,
    age = combined_person@age
  )
}
combine(me, hw)
```

- `class_any` matches any class
- `class_missing` for missing arguments

# Compatibility

## Compatibility with S3

- `class()` for S3 classes, `S7_class()` for S7 class constructor
- S7 properties are attributes (so old code that expects those will work)
- S7 can register methods for:
  - S7 class + S3 generic
  - S3 class + S7 generic
- S7 classes can inherit from S3 classes
- S3 classes can inherit from S7 classes

## Compatibility with S4

- S7 classes cannot inherit from S4 classes
- S4 classes can inherit from S7 classes
- S7 can register methods for:
  - S7 class + S4 generic
  - S4 class + S7 generic
- Out of scope: Both support class unions

::: notes
- I didn't try any S4 + S7 code, and there's a typo (I think) in that second bullet on the site (it says "S4 classes can inherit from S3 classes")
:::
