---
engine: knitr
title: Big picture
---

## Learning objectives:

- Become familiar with some metaprogramming principals and how they relate to each other
- Review vocabulary associated with metaprogramming

```{r}
library(rlang)
library(lobstr)
```


## Code is data

- **expression** - Captured code (*call*, *symbol*, *constant*, or *pairlist*)
- Use `rlang::expr()`[^1] to capture code directly

```{r}
expr(mean(x, na.rm = TRUE))
```

- Use `rlang::enexpr()` to capture code indirectly

```{r}
capture_it <- function(x) { # 'automatically quotes first argument'
  enexpr(x)
}
capture_it(a + b + c)
```

- 'Captured' code can be modified (like a list)!
    - First element is the function, next elements are the arguments

```{r}
f <- expr(f(x = 1, y = 2))
names(f)

ff <- fff <- f   # Create two copies

ff$z <- 3        # Add an argument to one
fff[[2]] <- NULL # Remove an argument from another

f
ff
fff
```

> More on this next week!

[^1]: Equivalent to `base::bquote()`

## Code is a tree

- **Abstract syntax tree** (AST) - Almost every language represents code as a tree
- Use `lobstr::ast()` to inspect these code trees

```{r}
ast(f1(f2(a, b), f3(1)))
ast(1 + 2 * 3)
```


## Code can generate code

- `rlang::call2()` creates function call

```{r}
call2("f", 1, 2, 3)
```

- Going backwards from the tree, can use functions to create calls

```{r}
call2("f1", call2("f2", "a", "b"), call2("f3", 1))
call2("+", 1, call2("*", 2, 3))
```

- `!!` bang-bang - **unquote operator**
    - inserts previously defined expressions into the current one

```{r}
xx <- expr(x + x)
yy <- expr(y + y)
expr(xx / yy)     # Nope!

expr(!!xx / !!yy) # Yup!
```

```{r}
cv <- function(var) {
  var <- enexpr(var)            # Get user's expression
  expr(sd(!!var) / mean(!!var)) # Insert user's expression
}

cv(x)
cv(x + y)
```

- Avoid `paste()` for building code
    - Problems with non-syntactic names and precedence among expressions

> "You might think this is an esoteric concern, but not worrying about it when generating SQL code in web applications led to SQL injection attacks that have collectively cost billions of dollars."

## Evaluation runs code

- **evaluate** - run/execute an expression
- need both expression and environment
- `eval()` uses current environment if not set
- manual evaluation means you can tweak the environment!

```{r}
xy <- expr(x + y)

eval(xy, env(x = 1, y = 10))
eval(xy, env(x = 2, y = 100))
```


## Customizing evaluations with functions
- Can also bind names to functions in supplied environment
- Allows overriding function behaviour
- This is how dplyr generates SQL for working with databases

For example...
```{r}
string_math <- function(x) {
  e <- env(
    caller_env(),
    `+` = function(x, y) paste(x, y),
    `*` = function(x, y) strrep(x, y)
  )

  eval(enexpr(x), e)
}

cohort <- 9
string_math("Hello" + "cohort" + cohort)
string_math(("dslc" + "is" + "awesome---") * cohort)
```


## Customizing evaluation with data

- Look for variables inside data frame
- **Data mask** - typically a data frame
- use `rlang::eval_tidy()` rather than `eval()`

```{r}
df <- data.frame(x = 1:5, y = sample(5))
eval_tidy(expr(x + y), df)
```

Catch user input with `enexpr()`...

```{r}
with2 <- function(df, expr) {
  eval_tidy(enexpr(expr), df)
}

with2(df, x + y)
```

But there's a bug!

- Evaluates in environment inside `with2()`, but the expression likely refers
  to objects in the Global environment
  
```{r}
with2 <- function(df, expr) {
  a <- 1000
  eval_tidy(enexpr(expr), df)
}

df <- data.frame(x = 1:3)
a <- 10
with2(df, x + a)
```

- Solved with Quosures...
  
## Quosures

- **Quosures** bundles expression with an environment
- Use `enquo()` instead of `enexpr()` (with `eval_tidy()`)

```{r}
with2 <- function(df, expr) {
  a <- 1000
  eval_tidy(enquo(expr), df)
}

df <- data.frame(x = 1:3)
a <- 10
with2(df, x + a)
```

> "Whenever you use a data mask, you must always use `enquo()` instead of `enexpr()`.

This comes back in Chapter 20.

### Which environment is bundled?
- The environment where the expression is created (i.e. the parent of where
  `enquo()` is called)

Here, the global environment

```{r}
with2 <- function(df, expr) {
  a <- 1000
  eq <- enquo(expr)
  message("with2() Parent/Calling environment: ")
  print(rlang::caller_env())
  message("with2() environment: ")
  print(rlang::current_env())
  message("Quosure details: ")
  print(eq)  # Print the details of the quosure
  eval_tidy(eq, df)
}

a <- 10000
df <- data.frame(x = 1:3)
with2(df, x + a)
```


Here, the `fun1()` environment
```{r}
fun1 <- function(df) {
  a <- 10
  message("fun1() Parent/Calling environment: ")
  print(rlang::caller_env())
  message("fun1() environment: ")
  print(rlang::current_env())
  with2(df, x + a)
}

a <- 10000
df <- data.frame(x = 1:3)
fun1(df)
```
