---
engine: knitr
title: Quasiquotation
---

## Learning objectives:

- What quasiquotation means
- Why it's important
- Learn some practical uses

```{r, message=FALSE}
library(rlang)
library(purrr)
```

## Introduction

Three pillars of *tidy* evaluation

   1. Quasiquotation
   2. Quosures (chapter 20)
   3. Data masks (Chapter 20)

**Quasiquotation = quotation + unquotation**

- **Quote.** Capture unevaluated expression... ("defuse")  
- **Unquote.** Evaluate selections of quoted expression! ("inject")
- Functions that use these features are said to use Non-standard evaluation (NSE)
- Note: related to Lisp macros, and also exists in other languages with Lisp heritage, e.g. Julia

> On it's own, Quasiquotation good for programming, but combined with other tools, 
> important for data analysis.

## Motivation

Simple *concrete* example:

`cement()` is a function that works like `paste()` but doesn't need need quotes

(Think of automatically adding 'quotes' to the arguments)

```{r}
cement <- function(...) {
  args <- ensyms(...)
  paste(purrr::map(args, as_string), collapse = " ")
}

cement(Good, morning, Hadley)
```

What if we wanted to use variables? What is an object and what should be quoted?

This is where 'unquoting' comes in!

```{r}
name <- "Bob"
cement(Good, afternoon, !!name) # Bang-bang!
```

## Vocabulary {-}

Can think of `cement()` and `paste()` as being 'mirror-images' of each other.

- `paste()` - define what to quote - **Evaluates** arguments
- `cement()` - define what to unquote - **Quotes** arguments

**Quoting function** similar to, but more precise than, **Non-standard evaluation (NSE)**

- Tidyverse functions - e.g., `dplyr::mutate()`, `tidyr::pivot_longer()`
- Base functions - e.g., `library()`, `subset()`, `with()`

**Quoting function** arguments cannot be evaluated outside of function:
```{r, error = TRUE}
cement(Good, afternoon, Cohort) # No problem
Good      # Error!
```

**Non-quoting (standard) function** arguments can be evaluated:
```{r}
paste("Good", "afternoon", "Cohort")
"Good"
```


## Quoting

**Capture expressions without evaluating them**

```{r, echo = FALSE}
data.frame(
  t = rep(c("One", "Many"), 3),
  Developer = c("`expr()`","`exprs()`", 
                "`quote()`", "`substitute()`",
                "", ""),
  User = c("`enexpr()`", "`enexprs()`", 
           "`alist()`", "`as.list(substitute(...()))`",
           "`ensym()`", "`ensyms()`"),
  type = c("Expression", "Expression", "R Base", "R Base", "Symbol", "Symbol")) |>
  dplyr::group_by(type) |>
  gt::gt() |>
  gt::tab_row_group(label = "R Base (Quotation)", rows = type == "R Base")|>
  gt::tab_row_group(label = "Symbol (Quasiquotation)", rows = type == "Symbol") |>
  gt::tab_row_group(label = "Expression (Quasiquotation)", rows = type == "Expression")|>
  gt::cols_label(t = "") |>
  gt::tab_options(row_group.font.weight = "bold") |>
  gt::tab_style(style = gt::cell_text(align = "center", weight = "bold"), 
                locations = gt::cells_column_labels()) |>
  gt::tab_style(style = gt::cell_borders(style = "hidden"), locations = gt::cells_body()) |>
  gt::tab_style(style = gt::cell_borders(sides = "top", style = "solid"),
                locations = gt::cells_body(rows = c(1, 3, 5))) |>
    gt::tab_style(style = gt::cell_borders(sides = "bottom", style = "solid"),
                locations = gt::cells_body(rows = c(2, 4))) |>
  gt::cols_align("center", columns = -1) |>
  gt::fmt_markdown() |>
  gt::cols_width(t ~ px(100))
```

- Non-base functions are from **rlang**
- **Developer** - From you, direct, fixed, interactive
- **User** - From the user, indirect, varying, programmatic

Also: 

- `bquote()` provides a limited form of quasiquotation
- `~`, the formula, is a quoting function (see [Section 20.3.4](https://adv-r.hadley.nz/evaluation.html#quosure-impl))

### `expr()` and `exprs()` {-}
```{r}
expr(x + y)
exprs(exp1 = x + y, exp2 = x * y)
```

### `enexpr()`^[`enexpr()` = **en**rich `expr()`] and `enexprs()` {-}
```{r}
f <- function(x) enexpr(x)
f(a + b + c)

f2 <- function(x, y) enexprs(exp1 = x, exp2 = y)
f2(x = a + b, y = c + d)
```

### `ensym()` and `ensyms()` {-}

- **[Remember](https://adv-r.hadley.nz/expressions.html#symbols):** Symbol represents the name of an object. Can only be length 1.
- These are stricter than `enexpr/s()`

```{r}
f <- function(x) ensym(x)
f(a)

f2 <- function(x, y) ensyms(sym1 = x, sym2 = y)
f2(x = a, y = "b")
```


## Unquoting

**Selectively evaluate parts of an expression**

- Merges ASTs with template
- 1 argument `!!` (**unquote**, **bang-bang**)
  - Unquoting a *function call* evaluates and returns results
  - Unquoting a *function (name)* replaces the function (alternatively use `call2()`)
- \>1 arguments `!!!` (**unquote-splice**, **bang-bang-bang**, **triple bang**)
- `!!` and `!!!` only work like this inside quoting function using rlang

### Basic unquoting {-}

**One argument**
```{r}
x <- expr(a + b)
y <- expr(c / d)
```

```{r, collapse = TRUE}
expr(f(x, y))      # No unquoting
expr(f(!!x, !!y))  # Unquoting
```

**Multiple arguments**
```{r}
z <- exprs(a + b, c + d)
w <- exprs(exp1 = a + b, exp2 = c + d)
```

```{r, collapse = TRUE}
expr(f(z))      # No unquoting
expr(f(!!!z))   # Unquoting
expr(f(!!!w))   # Unquoting when named
```


### Special usages or cases {-}

For example, get the AST of an expression
```{r, collapse = TRUE}
lobstr::ast(x)
lobstr::ast(!!x)
```


Unquote *function call*
```{r, collapse = TRUE}
expr(f(!!mean(c(100, 200, 300)), y))
```

Unquote *function*
```{r, collapse = TRUE}
f <- expr(sd)
expr((!!f)(x))
expr((!!f)(!!x + !!y))
```

## Non-quoting

Only `bquote()` provides a limited form of quasiquotation.

The rest of base selectively uses or does not use quoting (rather than unquoting). 

Four basic forms of quoting/non-quoting:

1. **Pair of functions** - Quoting and non-quoting
    - e.g., `$` (quoting) and `[[` (non-quoting)
2. **Pair of Arguments** - Quoting and non-quoting
    - e.g., `rm(...)` (quoting) and `rm(list = c(...))` (non-quoting)
3. **Arg to control quoting**
    - e.g., `library(rlang)` (quoting) and `library(pkg, character.only = TRUE)` (where `pkg <- "rlang"`)
4. **Quote if evaluation fails**
    - `help(var)` - Quote, show help for var
    - `help(var)` (where `var <- "mean"`) - No quote, show help for mean
    - `help(var)` (where `var <- 10`) - Quote fails, show help for var


## ... (dot-dot-dot) [When using ... with quoting]

- Sometimes need to supply an *arbitrary* list of expressions or arguments in a function (`...`)
- But need a way to use these when we don't necessarily have the names
- Remember `!!` and `!!!` only work with functions that use rlang
- Can use `list2(...)` to turn `...` into "tidy dots" which *can* be unquoted and spliced
- Require `list2()` if going to be passing or using `!!` or `!!!` in `...`
- `list2()` is a wrapper around `dots_list()` with the most common defaults

**No need for `list2()`**
```{r, collapse = TRUE}
d <- function(...) data.frame(list(...))
d(x = c(1:3), y = c(2, 4, 6))
```

**Require `list2()`**
```{r, collapse = TRUE, error = TRUE}
vars <- list(x = c(1:3), y = c(2, 4, 6))
d(!!!vars)
d2 <- function(...) data.frame(list2(...))
d2(!!!vars)
# Same result but x and y evaluated later
vars_expr <- exprs(x = c(1:3), y = c(2, 4, 6))
d2(!!!vars_expr)  
```

Getting argument names (symbols) from variables
```{r}
nm <- "z"
val <- letters[1:4]
d2(x = 1:4, !!nm := val)
```

## `exec()` [Making your own ...] {-}

What if your function doesn't have tidy dots?


Can't use `!!` or `:=` if doesn't support rlang or dynamic dots
```{r, collapse=TRUE, error = TRUE}
my_mean <- function(x, arg_name, arg_val) {
  mean(x, !!arg_name := arg_val)
}

my_mean(c(NA, 1:10), arg_name = "na.rm", arg_val = TRUE)     
```

Let's use the ... from `exec()`
```{r, eval = FALSE}
exec(.fn, ..., .env = caller_env())
```


```{r, collapse=TRUE}
my_mean <- function(x, arg_name, arg_val) {
  exec("mean", x, !!arg_name := arg_val)
}

my_mean(c(NA, 1:10), arg_name = "na.rm", arg_val = TRUE)     
```

Note that you do not unquote `arg_val`.
 
Also `exec` is useful for mapping over a list of functions:

```{r}
x <- c(runif(10), NA)
funs <- c("mean", "median", "sd")
purrr::map_dbl(funs, exec, x, na.rm = TRUE)
```

   
##  Base R `do.call` {-}

`do.call(what, args)`

- `what` is a function to call
- `args` is a list of arguments to pass to the function.

```{r, collapse = TRUE}
nrow(mtcars)
mtcars3 <- do.call("rbind", list(mtcars, mtcars, mtcars))
nrow(mtcars3)
```
 

### Exercise 19.5.5 #1 {-}

One way to implement `exec` is shown here: Describe how it works. What are the key ideas?

```{r}
exec_ <- function(f, ..., .env = caller_env()){
  args <- list2(...)
  do.call(f, args, envir  = .env)
}
```

## Case Studies (side note)

Sometimes you want to run a bunch of models, without having to copy/paste each one.

BUT, you also want the summary function to show the appropriate model call, 
not one with hidden variables (e.g., `lm(y ~ x, data = data)`). 

We can achieve this by building expressions and unquoting as needed:

```{r, collapse = TRUE}
library(purrr)

vars <- data.frame(x = c("hp", "hp"),
                   y = c("mpg", "cyl"))

x_sym <- syms(vars$x)
y_sym <- syms(vars$y)

formulae <- map2(x_sym, y_sym, \(x, y) expr(!!y ~ !!x))
formulae
models <- map(formulae, \(f) expr(lm(!!f, data = mtcars)))
summary(eval(models[[1]]))
```

As a function:
```{r, collapse = TRUE}
lm_df <- function(df, data) {
  x_sym <- map(df$x, as.symbol)
  y_sym <- map(df$y, as.symbol)
  data <- enexpr(data)
  
  formulae <- map2(x_sym, y_sym, \(x, y) expr(!!y ~ !!x))
  models <- map(formulae, \(f) expr(lm(!!f, !!data)))
  
  map(models, \(m) summary(eval(m)))
}

vars <- data.frame(x = c("hp", "hp"),
                   y = c("mpg", "cyl"))
lm_df(vars, data = mtcars)
```




## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/tbByqsRRvdE")`

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/IXE21pR8EJ0")`

### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/gxSpz6IePLg")`

### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/aniKrZrr4aU")`

### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/klcpEb5ZBSM")`

### Cohort 6

`r knitr::include_url("https://www.youtube.com/embed/OBodjc80y-E")`

<details>
<summary> Meeting chat log </summary>

```
01:02:07	Trevin:	Yeah, that was a great workshop
01:02:18	Trevin:	Glad they posted the resources online
01:06:39	Trevin:	Thank you!
```
</details>

### Cohort 7

`r knitr::include_url("https://www.youtube.com/embed/8LPw_VTBsmQ")`

<details>
<summary>Meeting chat log</summary>
```
00:50:48	Stone:	https://www.r-bloggers.com/2018/10/quasiquotation-in-r-via-bquote/
00:58:26	iPhone:	See ya next week!
```
</details>

`r knitr::include_url("https://www.youtube.com/embed/g77Jfl_xrXM")`

<details>
<summary>Meeting chat log</summary>
```
00:55:22	collinberke:	https://rlang.r-lib.org/reference/embrace-operator.html?q=enquo#under-the-hood
```
</details>
