---
engine: knitr
title: Improving performance
---

## Learning Objectives

0. Determine when to optimize code
1. Organize code for performance
2. Check for existing solutions
3. Create simple solutions
4. Vectorise code
5. Avoid code copying

```{r}
library(bench)
```

## Code that works, is correct, and is reproducible should be the first priority

*   Correct code generates a *good product*
*   Incorrect code generates a *bad product*
*   A good product is _more valuable_ than an bad product
*   People will use **and** wait for valuable products
*   People will not use incorrect products
*   Do not optimize code that doesn' yet generate a good product

## Organizing code

- Write a function for each approach
```{r}
mean1 <- function(x) mean(x)
mean2 <- function(x) sum(x) / length(x)
```
- Keep old functions that you've tried, even the failures
- Generate a representative test case
```{r}
x <- runif(1e5)
```
- Use `bench::mark` to compare the different versions (and include unit tests)
```{r}
bench::mark(
  mean1(x),
  mean2(x)
)[c("expression", "min", "median", "itr/sec", "n_gc")]
```

## Check for Existing Solution
- CRAN task views (http://cran.rstudio.com/web/views/)
- Reverse dependencies of Rcpp (https://cran.r-project.org/web/packages/Rcpp/)
- Talk to others!
  - Google (rseek)
  - Stackoverflow ([R])
  - https://community.rstudio.com/
  - DSLC community

## Do as little as possible
- use a function tailored to a more specific type of input or output, or to a more specific problem
  - `rowSums()`, `colSums()`, `rowMeans()`, and `colMeans()` are faster than equivalent invocations that use `apply()` because they are vectorised
  - `vapply()` is faster than `sapply()` because it pre-specifies the output type
  - `any(x == 10)` is much faster than `10 %in% x` because testing equality is simpler than testing set inclusion
- Some functions coerce their inputs into a specific type. If your input is not the right type, the function has to do extra work
  - e.g. `apply()` will always turn a dataframe into a matrix
- Other examples
  - `read.csv()`: specify known column types with `colClasses`. (Also consider
  switching to `readr::read_csv()` or `data.table::fread()` which are 
  considerably faster than `read.csv()`.)

  - `factor()`: specify known levels with `levels`.

  - `cut()`: don't generate labels with `labels = FALSE` if you don't need them,
  or, even better, use `findInterval()` as mentioned in the "see also" section
  of the documentation.
  
  - `unlist(x, use.names = FALSE)` is much faster than `unlist(x)`.

  - `interaction()`: if you only need combinations that exist in the data, use
  `drop = TRUE`.
  
## Avoiding Method Dispatch
```{r}
x <- runif(1e2)
bench::mark(
  mean(x),
  mean.default(x)
)[c("expression", "min", "median", "itr/sec", "n_gc")]
```

```{r}
x <- runif(1e2)
bench::mark(
  mean(x),
  mean.default(x),
  .Internal(mean(x))
)[c("expression", "min", "median", "itr/sec", "n_gc")]
```

```{r}
x <- runif(1e4)
bench::mark(
  mean(x),
  mean.default(x),
  .Internal(mean(x))
)[c("expression", "min", "median", "itr/sec", "n_gc")]
```

## Avoiding Input Coercion
- `as.data.frame()` is quite slow because it coerces each element into a data frame and then `rbind()`s them together
- instead, if you have a named list with vectors of equal length, you can directly transform it into a data frame

```{r}
quickdf <- function(l) {
  class(l) <- "data.frame"
  attr(l, "row.names") <- .set_row_names(length(l[[1]]))
  l
}
l <- lapply(1:26, function(i) runif(1e3))
names(l) <- letters
bench::mark(
  as.data.frame = as.data.frame(l),
  quick_df      = quickdf(l)
)[c("expression", "min", "median", "itr/sec", "n_gc")]
```

*Caveat!* This method is fast because it's dangerous!

## Vectorise
- vectorisation means finding the existing R function that is implemented in C and most closely applies to your problem
- Vectorised functions that apply to many scenarios
  - `rowSums()`, `colSums()`, `rowMeans()`, and `colMeans()`
  - Vectorised subsetting can lead to big improvements in speed
  - `cut()` and `findInterval()` for converting continuous variables to categorical
  - Be aware of vectorised functions like `cumsum()` and `diff()`
  - Matrix algebra is a general example of vectorisation

## Avoiding copies

- Whenever you use c(), append(), cbind(), rbind(), or paste() to create a bigger object, R must first allocate space for the new object and then copy the old object to its new home.

```{r}
random_string <- function() {
  paste(sample(letters, 50, replace = TRUE), collapse = "")
}
strings10 <- replicate(10, random_string())
strings100 <- replicate(100, random_string())
collapse <- function(xs) {
  out <- ""
  for (x in xs) {
    out <- paste0(out, x)
  }
  out
}
bench::mark(
  loop10  = collapse(strings10),
  loop100 = collapse(strings100),
  vec10   = paste(strings10, collapse = ""),
  vec100  = paste(strings100, collapse = ""),
  check = FALSE
)[c("expression", "min", "median", "itr/sec", "n_gc")]
```

## Case study: t-test

```{r}
m <- 1000
n <- 50
X <- matrix(rnorm(m * n, mean = 10, sd = 3), nrow = m)
grp <- rep(1:2, each = n / 2)
```

```{r}
# formula interface
system.time(
  for (i in 1:m) {
    t.test(X[i, ] ~ grp)$statistic
  }
)
# provide two vectors
system.time(
  for (i in 1:m) {
    t.test(X[i, grp == 1], X[i, grp == 2])$statistic
  }
)
```

Add functionality to save values

```{r}
compT <- function(i){
  t.test(X[i, grp == 1], X[i, grp == 2])$statistic
}
system.time(t1 <- purrr::map_dbl(1:m, compT))
```

If you look at the source code of `stats:::t.test.default()`, you’ll see that it does a lot more than just compute the t-statistic.

```{r}
# Do less work
my_t <- function(x, grp) {
  t_stat <- function(x) {
    m <- mean(x)
    n <- length(x)
    var <- sum((x - m) ^ 2) / (n - 1)
    list(m = m, n = n, var = var)
  }
  g1 <- t_stat(x[grp == 1])
  g2 <- t_stat(x[grp == 2])
  se_total <- sqrt(g1$var / g1$n + g2$var / g2$n)
  (g1$m - g2$m) / se_total
}
system.time(t2 <- purrr::map_dbl(1:m, ~ my_t(X[.,], grp)))
stopifnot(all.equal(t1, t2))
```

This gives us a six-fold speed improvement!

```{r}
# Vectorise it
rowtstat <- function(X, grp){
  t_stat <- function(X) {
    m <- rowMeans(X)
    n <- ncol(X)
    var <- rowSums((X - m) ^ 2) / (n - 1)
    list(m = m, n = n, var = var)
  }
  g1 <- t_stat(X[, grp == 1])
  g2 <- t_stat(X[, grp == 2])
  se_total <- sqrt(g1$var / g1$n + g2$var / g2$n)
  (g1$m - g2$m) / se_total
}
system.time(t3 <- rowtstat(X, grp))
stopifnot(all.equal(t1, t3))
```

1000 times faster than when we started!

## Other techniques
* [Read R blogs](http://www.r-bloggers.com/) to see what performance
  problems other people have struggled with, and how they have made their
  code faster.

* Read other R programming books, like The Art of R Programming or Patrick Burns'
  [_R Inferno_](http://www.burns-stat.com/documents/books/the-r-inferno/) to
  learn about common traps.

* Take an algorithms and data structure course to learn some
  well known ways of tackling certain classes of problems. I have heard
  good things about Princeton's
  [Algorithms course](https://www.coursera.org/course/algs4partI) offered on
  Coursera.
  
* Learn how to parallelise your code. Two places to start are
  Parallel R and Parallel Computing for Data Science

* Read general books about optimisation like Mature optimisation
  or the Pragmatic Programmer
  
* Read more R code. StackOverflow, R Mailing List, DSLC, GitHub, etc.
