---
engine: knitr
title: Rewriting R code in C++/Rust
---

# _Interfaces to other software are part of R_  

--- Chambers JM (2016). Extending R.

## Learning objectives:

-   Why rewrite R code in C++/Rust
-   Compare C++/Rust function and package development
-   Run C++/Rust in R
-   Create and call C++/Rust functions in R
-   Evaluate function performance for R/C/C++/Rust
-   Compare compiler error messages between C++ and Rust

```{r}
library(Rcpp)
library(rextendr)
```

## Even after optimizing R code, it still might not be optimized enough

::: panel-tabset

### C++

-   R's for loops are much slower compared to C++

    - Can't vectorize in R when loops depend on subsequent iterations

-   R function calls are much slower compared to C++

    - Recursive functions involving millions of function calls

-   R does not include many data structures compared to C++

    - C++ Standard Template Library (STL) for efficiency, correctness, and maintainability

### Rust

-   R cannot exploit parallelism as effectively as Rust

    - Rust memory is safer (ownership model with borrow checking) and so lacks GIL or run time locks

-   R object memory is borrowed via C/C++ unlike Rust

    - Rust owns objects and code compiles directly to machine code

-   R or C code debugging is more challenging than Rust

    - Rust has no GC so lacks memory bugs and has great error messages comparable with the `{tidyverse}`

:::

## C++ and Rust code are used via packages and offer similar implementation approaches

::: panel-tabset

### C++

::: panel-tabset

#### `Rcpp::evalCpp`

```{r}
Rcpp::evalCpp('NA_INTEGER + 1')
```

#### `Rcpp::cppFunction`

```{r}
Rcpp::cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')
add
add(1,2,3)
```

#### `Rcpp::sourceCpp`

```c
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;

  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}
```

#### `usethis::use_rcpp`

-   Creates infrastructure for using C++ within R packages

:::

### Rust

::: panel-tabset

#### `rextendr::rust_function`

```{r}
rextendr::rust_function("fn add(a:f64, b:f64) -> f64 { a + b }")
add(2.5, 4.7)
```

#### `rextendr::rust_source`

```{r}
code <- '
#[extendr]
fn main() {
  println!("Hello, world!");
}
'

rextendr::rust_source(
  code = code
)

main()
```

#### `rextendr::use_extendr`

```r
# create a new R package
usethis::create_package("helloextendr")

# Use extendr
rextendr::use_extendr()

# Document and build the package
rextendr::document()

# run hello_world()
hello_world()
#> [1] "Hello, world!"
```

-   Newer Rust toolchain and R version is required
-   Using VSCode/Positron is strongly recommended with `rust-analyzer` extension

:::

:::

## Reading C++ and Rust code is similar to reading R code

:::panel-tabset

### Scalar functions

::: panel-tabset

#### R

```{r}
signR <- function(x) {
  if (x > 0) {
    1
  } else if (x == 0) {
    0
  } else {
    -1
  }
}
```

#### C++

```{r}
Rcpp::cppFunction('int signC(double x) {
  if (x > 0) {
    return 1;
  } else if (x == 0) {
    return 0;
  } else {
    return -1;
  }
}') |> system.time()
```

#### Rust

```{r}
rextendr::rust_function("
fn signRust(x: f64) -> i32 {
    if x > 0.0 {
        1
    } else if x == 0.0 {
        0
    } else {
        -1
    }
}
") |> system.time()
```

:::

### Vector functions

::: panel-tabset

#### R

```{r}
sign

signR <- function(x) {
  as.integer(vapply(x,sign,double(1)))
}
```

#### C++

```{r}
Rcpp::cppFunction("
NumericVector signC(NumericVector x) {
  int n = x.size();
  NumericVector out(n);

  for (int i = 0; i < n; i++) {
    if (x[i] > 0.0) {
      out[i] = 1;
    } else if (x[i] == 0.0) {
      out[i] = 0;
    } else {
      out[i] = -1;
    }
  }

  return out;
}") |> system.time()

Rcpp::cppFunction("
NumericVector signCfaster(NumericVector x) {
  int n = x.size();
  NumericVector out(n);

  for (int i = 0; i < n; ++i) {
    double v = x[i];
    out[i] = (v > 0) ? 1.0 : (v == 0 ? 0.0 : -1.0);
  }

  return out;

}") |> system.time()
```

#### Rust

```{r}
rextendr::rust_function("
fn signRust(x: Vec<f64>) -> Vec<Rint> {
    x.into_iter()
        .map(|v| {
            if v > 0.0 {
                Rint::from(1)
            } else if v == 0.0 {
                Rint::from(0)
            } else {
                Rint::from(-1)
            }
        })
        .collect()
}
") |> system.time()

code <- '

use rayon::prelude::*;

#[extendr]
fn signRustPar(x: Vec<f64>) -> Vec<Rint> {
    x.into_par_iter()
        .map(|v| {
            if v > 0.0 {
                Rint::from(1)
            } else if v == 0.0 {
                Rint::from(0)
            } else {
                Rint::from(-1)
            }
        })
        .collect()
}
'

rextendr::rust_source(
  code = code,
  dependencies = list(rayon = "1")
) |> 
  system.time()
```

:::

:::

## The C optimized `sign` is most performant, then C++ re-implementation is faster but Rust uses less memory

```{r}
vec <- rnorm(1e7,sd = 50)
bench::mark(
  signR(vec),
  sign(vec),
  signCfaster(vec),
  signC(vec),
  signRustPar(vec),
  signRust(vec),
  relative = TRUE
)
```

## Rust has more readable error messages than C++

::: panel-tabset

### C++

```{r, error = TRUE}
Rcpp::cppFunction("
NumericVector signCfaster(NumericVector x) {
  int n = x.size();
  NumericVector out(n);

  for (int i = 0; i < n; ++i) {
    out[i] = (v > 0) ? 1.0 : (v == 0 ? 0.0 : -1.0);
  }

  return out;

}")
```

### Rust

```{r,error = TRUE}
rextendr::rust_function('
fn signRustSlice(x: Robj) -> Robj {
    // Borrow numeric slice from R (no copy)
    let slice = x.as_real_slice().expect("Expected numeric vector");

    let n = slice.len();
    let mut out = Doubles::new(n);

    // Unsafe block gives direct mutable slice (no bounds checks)
    let out_slice = unsafe { out.as_mut_slice() };

    for i in 0..n {
        let v = slice[i];
        out_slice[i] = if v.is_nan() {
            f64::NAN
        } else if v > 0.0 {
            1.0
        } else if v == 0.0 {
            0.0
        } else {
            -1.0
        };
    }

    out.into()
}
')
```

:::

## References

-   Rust
    -   [crates.io: Rust package manager](https://crates.io)
    -   [`{cargo}` book](https://doc.rust-lang.org/cargo/)
    -   [`{rextendr} on CRAN`](https://cran.r-project.org/web/packages/rextendr/index.html)
    -   [extendr_api crate documentation](https://docs.rs/extendr-api/latest/extendr_api/)
-   C++
    -   [vcpkg C++ package manager](https://vcpkg.io/en/)
    -   [`{Rcpp}` on CRAN](https://cran.r-project.org/web/packages/Rcpp/index.html)
    -   [Rcpp for everyone](https://doc.rust-lang.org/cargo/)
-   [Rust and C++ comparison gitbook](https://locka99.gitbooks.io/a-guide-to-porting-c-to-rust/content/)